# Solution

## Task 1

(a)
```{r, eval=TRUE}
X <- physM$Weight
Y <- physM$Power1

c(var(X), var(Y), cov(X, Y))
```

(b)
```{r, eval=TRUE}
cov(X, Y)/sqrt(var(X)*var(Y))
cor.test(physM$Weight, physM$Power1)$estimate
```

(c)
```{r, eval=TRUE}
Sxx <- sum((X-mean(X))^2)
Syy <- sum((Y-mean(Y))^2)
Sxy <- sum((X-mean(X))*(Y-mean(Y)))

c(Sxx, Syy, Sxy)
```

(d)
```{r, eval=TRUE}
Sxy/sqrt(Sxx*Syy)
```

(e)
```{r, eval=TRUE}
X <- physF$Power1
Y <- physF$Weight

cor.test(X, Y)

cov(X, Y)/sqrt(var(X)*var(Y))

Sxx <- sum((X-mean(X))^2)
Syy <- sum((Y-mean(Y))^2)
Sxy <- sum((X-mean(X))*(Y-mean(Y)))
Sxy / sqrt(Sxx * Syy)

```

## Task 2

(a) Producing a scatterplot
```{r}
plot(NumDrowned ~ NumFilms, data = Cage, xlab = "Number of Nicholas Cage films released in a year", ylab = "Number of people who drowned falling into pools that year")
```

(b) Correlation test
```{r, eval=TRUE}
cor.test(Cage$NumDrowned, Cage$NumFilms)
```

The sample correlation coefficient is 0.666, indicating a moderate positive linear relationship between **NumDrowned** and **NumFilms**. 

The $p$-value is less than 0.05. Therefore, we reject the null hypothesis and conclude that the population correlation $\rho$ is statistically significantly different from 0. 
