---
title: |
    | S2Y Lab 2
    | Understanding correlation and further exploring relationships
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
biblio-style: apalike
link-citations: yes
---

```{r setup, include = FALSE}
library(webexercises)
library(tinytex)
library(dplyr)
library(knitr)
library(ggplot2)
library(tidyverse)
library(PASWR)

phys <- read.csv("phys1.csv")

knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```


```{r include=FALSE, echo=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

# Welcome to S2Y Lab 2

Intended Learning Outcomes:

- calculate and interpret the sample correlation coefficient;

- perform hypothesis tests on the population correlation and interpret the decision;

- calculate least squares estimates of model parameters using vector-matrix formulation.

## Introduction

In the lectures we learned how to assess the strength of a linear relationship between random variables using the correlation coefficient.  The population correlation is a measure of the magnitude of the strength of the relationship between two random variables $X$ and $Y$, and is defined as

\begin{equation}
\rho(X,Y) = \frac{\text{Cov}(X,Y)}{\sqrt{\text{Var}(X)\text{Var}(Y)}},
(\#eq:eq1)
\end{equation}

and can be estimated by replacing each of $\text{Cov}(X,Y)$, $\text{Var}(X)$ and $\text{Var}(Y)$ by their unbiased estimators to give

\begin{equation}
r = \frac{S_{xy}}{\sqrt{S_{xx}S_{yy}}}=\frac{\sum^{n}_{i=1}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum^{n}_{i=1}(x_i-\bar{x})^2(y_i-\bar{y})^2}},
(\#eq:eq2)
\end{equation}

the **sample correlation coefficient** ($-1 \le r \le 1$).

Given a sample of data, we can assess the statistical significance of the observed correlations between
variables in the wider population. To do this we perform a hypothesis test (more on this in Example 1).

